---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%",
  dev = 'svg'
)
```

# 🦋 metamorphr

<!-- badges: start -->

<!-- badges: end -->

metamorphr provides a set of wrapper functions to make working with metabolomics data more fun. All data is stored in one [tidy](https://cran.r-project.org/web/packages/tidyr/vignettes/tidy-data.html) [tibble](https://tibble.tidyverse.org/) which facilitates integration with [tidyverse](https://www.tidyverse.org/) packages like [dplyr](https://dplyr.tidyverse.org/) and [ggplot2](https://ggplot2.tidyverse.org/). Also, functions in metamorphr are [pipe](https://magrittr.tidyverse.org/index.html)-friendly.

## 📥 Installation

You can install the development version of metamorphr from [GitHub](https://github.com/) with:

``` r
# install.packages("pak")
pak::pak("yasche/metamorphr")
```

## 🧹 Introducing a (somewhat) tidy data format

A prerequisite for frictionless working with [tidyverse](https://www.tidyverse.org/) packages is, that data is stored in a [tidy](https://cran.r-project.org/web/packages/tidyr/vignettes/tidy-data.html) format. Metabolomics feature tables usually do not fulfill the criteria for tidy data. They are normally a data frame or a matrix with features in rows and samples in columns or vice versa. This can lead to frustration when applying [dplyr](https://dplyr.tidyverse.org/) or [ggplot2](https://ggplot2.tidyverse.org/) functions. metamorphr provides convenience functions to read feature tables into a (somewhat) tidy format and add sample metadata.

Consider this small feature table saved as `my_featuretable.csv` with a label column, `label`, 2 columns that hold additional feature metadata, `feature_metadata1` and `feature_metadata2` (*e.g.*, retention time and *m/z* or some identifier like HMDB or KEGG) and 2 samples, `sample1` and `sample2`. It further holds the the measured intensity (or area) of 2 features, `feature1` and `feature2` for both samples:

```{r create-example-featuretable, include=FALSE}
library(tibble)
library(readr)

ftable_string <- "label,feature_metadata1,feature_metadata2,sample1,sample2
feature1,metadata1_1,metadata2_1,1,2
feature2,metadata2_1,metadata2_2,3,4
"
```

```{r example-featuretable-1, echo=FALSE, warning=FALSE, message=FALSE}
readr::read_csv(ftable_string) %>%
  knitr::kable()
```

The convenience function `read_featuretable()` can read a delimited file (like csv or tsv), a connection or literal data and transform it to a (somewhat) tidy tibble:

```{r example-featuretable-3, eval=FALSE}
library(metamorphr)

read_featuretable("my_featuretable.csv", label_col = 1, metadata_cols = c(2,3))
```

```{r example-featuretable-2, echo=FALSE, warning=FALSE, message=FALSE}
library(metamorphr)

read_featuretable(ftable_string, label_col = 1, metadata_cols = c(2,3)) %>%
  knitr::kable()
```

As you can see, the function reads the file and performs some transformation. The first column is a unique identifier (`UID`) for the features. Column 2 (`Feature`) holds the labels from the original feature table. The third column (`Sample`) specifies the sample and the fourth (`Intensity`) holds the measured intensity (or area). Columns 5 and 6 (`feature_metadata1` and `feature_metadata2`) hold additional feature metadata. These columns are redundant because they can be predicted from the `UID` column. It is up to the user to drop them or leave them in. They don't do any harm and might even be useful for downstream filtering or plotting.

## 📑 Examples

### 📊 Easy dplyr and ggplot2 integration

Easily plot the distribution of all intensities across samples

```{r example-dplyr-1, warning=FALSE, message=FALSE}
library(dplyr)
library(ggplot2)

toy_metaboscape %>%
  dplyr::left_join(toy_metaboscape_metadata, by = "Sample") %>%
  ggplot2::ggplot(ggplot2::aes(Sample, Intensity, color = Group)) +
    ggplot2::geom_boxplot() +
    theme_bw()
```

... or compare the intensity of specific features across groups

```{r example-dplyr-2, warning=FALSE}
toy_metaboscape %>%
  dplyr::left_join(toy_metaboscape_metadata, by = "Sample") %>%
  dplyr::filter(Name %in% c("Arachidonic acid", "ADP", "NADPH")) %>%
  dplyr::filter(Group %in% c("control", "treatment")) %>%
  ggplot2::ggplot(ggplot2::aes(Group, Intensity, color = Group)) +
    ggplot2::geom_boxplot() +
    ggplot2::facet_wrap(~Name) +
    ggplot2::theme_bw()
```
