---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%",
  dev = 'svg'
)
```

# 🦋 metamorphr

<!-- badges: start -->

<!-- badges: end -->

metamorphr provides a set of wrapper functions to make working with metabolomics data more fun. All data is stored in one [tidy](https://cran.r-project.org/web/packages/tidyr/vignettes/tidy-data.html) [tibble](https://tibble.tidyverse.org/) which facilitates integration with [tidyverse](https://www.tidyverse.org/) packages like [dplyr](https://dplyr.tidyverse.org/) and [ggplot2](https://ggplot2.tidyverse.org/). Also, functions in metamorphr are [pipe](https://magrittr.tidyverse.org/index.html)-friendly.

## 📥 Installation

You can install the development version of metamorphr from [GitHub](https://github.com/) with:

``` r
# install.packages("pak")
pak::pak("yasche/metamorphr")
```

## 🧹 Introducing a (somewhat) tidy data format

**TL;DR** Skip to [📑 Examples](#-examples).

A prerequisite for frictionless working with [tidyverse](https://www.tidyverse.org/) packages is that data is stored in a [tidy](https://cran.r-project.org/web/packages/tidyr/vignettes/tidy-data.html) format. Metabolomics feature tables usually do not fulfill the criteria for tidy data. They are normally a data frame or a matrix with features in rows and samples in columns or vice versa. This can lead to frustration when applying [dplyr](https://dplyr.tidyverse.org/) or [ggplot2](https://ggplot2.tidyverse.org/) functions. metamorphr provides convenience functions to read feature tables into a (somewhat) tidy format and add sample metadata.

Consider this small feature table saved under `my_featuretable.csv` with a label column, `label`, 2 columns that hold additional feature metadata, `feature_metadata1` and `feature_metadata2` (*e.g.*, retention time and *m/z* or some identifier like HMDB or KEGG) and 2 samples, `sample1` and `sample2`. It further holds the measured intensity (or area) of 2 features, `feature1` and `feature2` for both samples:

```{r create-example-featuretable, include=FALSE}
library(tibble)
library(readr)

ftable_string <- "label,feature_metadata1,feature_metadata2,sample1,sample2
feature1,metadata1_1,metadata2_1,1,2
feature2,metadata2_1,metadata2_2,3,4
"
```

```{r example-featuretable-1, echo=FALSE, warning=FALSE, message=FALSE}
readr::read_csv(ftable_string) %>%
  knitr::kable()
```

The convenience function `read_featuretable()` can read a delimited file (like csv or tsv), a connection or literal data and transform it to a (somewhat) tidy tibble:

```{r example-featuretable-2, eval=FALSE}
library(metamorphr)

my_featuretable <- read_featuretable("my_featuretable.csv", label_col = 1, metadata_cols = c(2,3))
```

The tibble `my_featuretable` looks like this:

```{r example-featuretable-3, echo=FALSE, warning=FALSE, message=FALSE}
library(metamorphr)

my_featuretable <- read_featuretable(ftable_string, label_col = 1, metadata_cols = c(2,3))
knitr::kable(my_featuretable)
```

As you can see, the function reads the file and performs some transformation. The first column is a unique identifier (`UID`) for the features. Column 2 (`Feature`) holds the labels from the original feature table. The third column (`Sample`) specifies the sample and the fourth (`Intensity`) holds the measured intensity (or area). Columns 5 and 6 (`feature_metadata1` and `feature_metadata2`) hold additional feature metadata. These columns are redundant because they can be predicted from the `UID` column. It is up to the user to drop them or leave them in. They don't do any harm and might even be useful for downstream filtering or plotting.

Sample metadata like (treatment) groups or backgrounds can easily be added with the 2 functions `create_metadata_skeleton()` and `join_metadata`. `create_metadata_skeleton()` creates an empty tibble for you to populate with metadata:

```{r example-create-metadata-1}
my_metadata <- my_featuretable %>%
  create_metadata_skeleton()
```

This is the resulting `my_metadata` tibble:

```{r example-create-metadata-2, echo=FALSE, warning=FALSE, message=FALSE}
knitr::kable(my_metadata)
```

By default it has 5 columns and as many rows as there are unique samples in the underlying featuretable (2 in this case). The first column, `Sample`, holds the samples. It is already populated with the names of the samples. The `Group` column is empty and can store grouping information, for example information about treatments or backgrounds. The third column, `Replicate` is important if there are multiple technical replicates, for example if a sample was injected multiple times or workup was performed with the same starting material multiple times. Then, the same samples should have the same `Replicate` number (or character) and `Group` name. `Batch` specifies the batch. It is only applicable if the sample or measurement was performed in multiple batches. Finally, `Factor` can hold a numeric factor like dry weight or protein concentration for sample-specific normalization.

Below is an example for how to populate the metadata table in R. Alternatively, it can also be saved to disk (e.g., with `readr::write_csv()`), edited manually and read again (e.g., with `readr::read_csv()`).

```{r example-create-metadata-3}
my_metadata <- my_metadata %>%
  dplyr::mutate(
    Group = c("treatment", "control"),
    Replicate = c(1L, 1L),
    Batch = c(1L, 1L),
    Factor = c(0.976, 1.035)
  )
```

Finally, `my_featuretable` and `my_metadata` can be joined via `join_metadata()`.

```{r example-create-metadata-4}
my_featuretable <- my_featuretable %>%
  join_metadata(my_metadata)
```

```{r example-create-metadata-5, echo=FALSE, warning=FALSE, message=FALSE}
knitr::kable(my_featuretable)
```

## 📑 Examples

### 📊 Easy dplyr and ggplot2 integration

Easily plot the distribution of all intensities across samples

```{r example-dplyr-1, warning=FALSE, message=FALSE}
library(dplyr)
library(ggplot2)

toy_metaboscape %>%
  join_metadata(toy_metaboscape_metadata) %>%
  ggplot2::ggplot(ggplot2::aes(Sample, Intensity, color = Group)) +
    ggplot2::geom_boxplot() +
    theme_bw()
```

... or compare the intensity of specific features across groups

```{r example-dplyr-2, warning=FALSE}
toy_metaboscape %>%
  join_metadata(toy_metaboscape_metadata) %>%
  dplyr::filter(Name %in% c("Arachidonic acid", "ADP", "NADPH")) %>%
  dplyr::filter(Group %in% c("control", "treatment")) %>%
  ggplot2::ggplot(ggplot2::aes(Group, Intensity, color = Group)) +
    ggplot2::geom_boxplot() +
    ggplot2::facet_wrap(~Name) +
    ggplot2::theme_bw()
```
