---
title: "Conjugate Screening"
output: 
  rmarkdown::html_vignette:
    toc: true
vignette: >
  %\VignetteIndexEntry{Conjugate Screening}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE, warning = FALSE}
safe_read_csv <- purrr::safely(readr::read_csv)
safe_read_lines <- purrr::safely(readr::read_lines)

#check if all necessary files can be read
all_available <- all(is.null(safe_read_csv("https://github.com/yasche/metamorphr-data/raw/refs/heads/main/RP18/pos/MetaboScape/menadione/menadione.csv", show_col_types = FALSE)$error),
                     is.null(safe_read_csv("https://github.com/yasche/metamorphr-data/raw/refs/heads/main/RP18/pos/MetaboScape/menadione/menadione_metadata.csv", show_col_types = FALSE)$error),
                     is.null(safe_read_lines("https://github.com/yasche/metamorphr-data/raw/refs/heads/main/RP18/pos/MetaboScape/menadione/menadione.gnps.mgf", n_max = 1)$error))

knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  error = TRUE,
  eval = all_available
)
```

```{r eval = !all_available, echo = FALSE, comment = NA}
message("Can't download example files from GitHub. Code in this vignette will not be evaluated.")
```

```{r lib}
library(metamorphr)
```

This vignette explains how to use the metamorphr package to search a featuretable for characteristic MS/MS fragments and neutral losses.

## Introduction

Glutathion (GSH) conjugates of xenobiotics usually result in characteristic MS/MS fragments and neutral losses (Brink et al., 2014). This vignette shows you two approaches, one relying on MS/MS fragments and one relying on neutral losses, to screen a metabolomics data set with associated MS/MS spectra for GSH conjugates. If you want to learn more about drug metabolism you can visit the Wikipedia page (<https://en.wikipedia.org/wiki/Drug_metabolism>). The data set was obtained by treating HepG2 cells with the redox cycler menadione and measuring the samples in ESI positive mode.

## Prepare the data set

This vignette uses data from a separate [GitHub repository](https://github.com/yasche/metamorphr-data). More information about the data can be found there.

First, the featuretable, the sample metadata, and the MS/MS spectra have to be imported and joined together. The metamorphr package provides several convenience functions to help with these tasks. To read the sample metadata in the form of a .csv file, `readr::read_csv` can be used.

```{r read-files, warning=FALSE}
menadione_ft <- read_featuretable(
  "https://github.com/yasche/metamorphr-data/raw/refs/heads/main/RP18/pos/MetaboScape/menadione/menadione.csv", 
  label_col = 3, 
  metadata_cols = 1:15)

menadione_metadata <- readr::read_csv(
  "https://github.com/yasche/metamorphr-data/raw/refs/heads/main/RP18/pos/MetaboScape/menadione/menadione_metadata.csv", 
  show_col_types = FALSE)

menadione_msn <- read_mgf("https://github.com/yasche/metamorphr-data/raw/refs/heads/main/RP18/pos/MetaboScape/menadione/menadione.gnps.mgf",
                          show_progress = F)
```

After the files have been imported successfully, they can be joined together. First the sample metadata is joined using the convenience function `join_metadata`. The MS/MS spectra can simply be added via `dplyr::left_join`.

```{r join-files}
menadione_ft <- menadione_ft %>%
  join_metadata(menadione_metadata) %>%
  dplyr::left_join(menadione_msn, by = "FEATURE_ID")

head(menadione_ft)
```

## Find potential glutathion conjugates

To find potential GSH conjugates, two approaches can be used: (i) search for specific MS/MS spectra and (ii) search for specific neutral losses. Functions for both approaches are implemented in the metamorphr package and are shown below. Results can be easily visualized using the 'ggplot2' package. Characteristic fragments and neutral losses are described in Brink et al., 2014. 

### Some initial filtering

To reduce the search space and remove potential false-positives, some initial filtering can be done using one or more of the several filter functions provided by the metamorphr package. First, QC samples are removed using `dplyr::filter` as they are not needed for further analysis. In this scenario, we are only interested in features that are solely present in the treated samples, so `filter_blank` can be used prior to searching for specific fragments or neutral losses. Additionally, we are only interested in features that are found reproducibly across samples, so we also use `filter_grouped_mv`. Finally, blank samples are removed, again using `dplyr::filter`.

```{r init-filter}
menadione_ft <- menadione_ft %>%
  dplyr::filter(Group != "QC") %>%
  filter_grouped_mv(min_found = 0.75)  %>%
  filter_blank(blank_samples = c("blank", "control"),
               blank_as_group = T,
               group_column = Group,
               min_frac = 3) %>%
  dplyr::filter(Group != "blank")
```

### Characteristic MS/MS fragments

Searching for characteristic MS/MS fragments can be done using the `filter_msn` function.

```{r filter-ms-ms}
fragments <- c(308.0911,
               179.0485,
               177.0328,
               162.0219,
               130.0499)

menadione_ft %>%
  filter_msn(fragments = fragments, 
             min_found = 2, 
             tolerance = 5, 
             tolerance_type = "ppm", 
             show_progress = F) %>%
  ggplot2::ggplot(ggplot2::aes(Group, Intensity)) +
    ggplot2::geom_point(position = "jitter") +
    ggplot2::facet_wrap(~ Feature) +
    ggplot2::ylim(c(0, NA))
```

Searching for characteristic MS/MS fragments yields only one potential GSH conjugate: A feature with an exact mass of 511.12575 Da.

### Characteristic neutral losses

Prior the searching for specific neutral losses via the `filter_neutral_loss` function, neutral loss spectra must be calculated using `calc_neutral_loss`. The usage of both functions is shown below.

```{r filter-losses}
losses <- c(75.0320,
            129.0426,
            232.0696,
            146.0692,
            249.0961,
            273.0961,
            275.1117,
            305.0682,
            307.0838)

menadione_ft %>%
  calc_neutral_loss(m_z_col = PEPMASS) %>%
  filter_neutral_loss(losses = losses, 
                      min_found = 4, 
                      tolerance = 10, 
                      tolerance_type = "ppm",
                      show_progress = F) %>%
  ggplot2::ggplot(ggplot2::aes(Group, Intensity)) +
    ggplot2::geom_point(position = "jitter") +
    ggplot2::facet_wrap(~ Feature) +
    ggplot2::ylim(c(0, NA))
```

This approach yields two potential GSH conjugates: One with an exact mass of 477.12018 and the one with an exact mass of 511.12575 which was also found using the MS/MS fragments-based approach above.

Interestingly, menadione has been reported to form a GSH conjugate, thiodione, with a calculated exact mass of 477.12058 Da (Mauzeroll et al., 2004). The difference between the measured exact mass of one potential conjugate and the calculated mass is just 0.0004 Da or 0.86 ppm.

## References

Brink, A., Fontaine, F., Marschmann, M., Steinhuber, B., Cece, E. N., Zamora, I., & Pähler, A. (2014). Post-acquisition analysis of untargeted accurate mass quadrupole time-of-flight MS E data for multiple collision-induced neutral losses and fragment ions of glutathione conjugates: Neutral losses and fragment ion analysis of glutathione conjugates. *Rapid Communications in Mass Spectrometry, 28*(24), 2695–2703. <https://doi.org/10.1002/rcm.7062>

Mauzeroll, J., Bard, A. J., Owhadian, O., & Monks, T. J. (2004). Menadione metabolism to thiodione in hepatoblastoma by scanning electrochemical microscopy. *Proceedings of the National Academy of Sciences, 101*(51), 17582–17587. <https://doi.org/10.1073/pnas.0407613101>
