---
title: "metamorphr"
output: 
  rmarkdown::html_vignette:
    toc: true
vignette: >
  %\VignetteIndexEntry{metamorphr}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE, warning = FALSE}
safe_read_csv <- purrr::safely(readr::read_csv)
#check if all necessary files can be read

all_available <- all(is.null(safe_read_csv("https://raw.githubusercontent.com/yasche/metamorphr-data/main/RP18/pos/MetaboScape/mevastatin/mevastatin.csv", show_col_types = FALSE)$error),
                     is.null(safe_read_csv("https://raw.githubusercontent.com/yasche/metamorphr-data/main/RP18/pos/MetaboScape/mevastatin/mevastatin_metadata.csv", show_col_types = FALSE)$error))

knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  error = TRUE,
  eval = all_available
)
```

```{r eval = !all_available, echo = FALSE, comment = NA}
message("Can't download example files from GitHub. Code in this vignette will not be evaluated.")
```

```{r lib}
library(metamorphr)
```

This vignette shows you how to work with metabolomics data using the metamorphr package. It is intended as a quick overview and the choices made (e.g., the algorithm for missing value imputation or normalization) may not be the best for the specific data set.

## Introduction

Metabolomics data preparation usually involves several steps: Feature tables are filtered to only contain 'significant' features, missing values are imputed, intensities are normalized, and, depending on the downstream analysis, data might also be scaled or transformed. The metamorphr package was developed to make these steps more streamlined and easier to perform. An example workflow is shown below. The goal of the workflow is to create a volcano plot that compares cells treated with the HMG-CoA reductase inhibitor mevastatin to cells treated with a vehicle control.

## The whole workflow

For a quick reference, the workflow is shown as a whole below. The individual steps are explained in detail further down.

*The metamorphr package was developed to be used with a [pipe operator](https://r4ds.hadley.nz/data-transform.html#sec-the-pipe) and this vignette makes extensive use of it. Make sure that you are familiar with the concept of the pipe before reading on. Also, this vignette uses the [magrittr pipe](https://magrittr.tidyverse.org/reference/pipe.html) instead of the [native pipe](https://r4ds.hadley.nz/data-transform.html#sec-the-pipe) to make it compatible with older versions of R. You should be able to use the native pipe as a drop-in replacement for the magrittr pipe, if you wish.*

```{r whole-thing, warning=FALSE}
mevastatin_ft <- read_featuretable(
  "https://raw.githubusercontent.com/yasche/metamorphr-data/main/RP18/pos/MetaboScape/mevastatin/mevastatin.csv", 
  label_col = 3, 
  metadata_cols = 1:15)

mevastatin_metadata <- readr::read_csv(
  "https://raw.githubusercontent.com/yasche/metamorphr-data/main/RP18/pos/MetaboScape/mevastatin/mevastatin_metadata.csv", 
  show_col_types = FALSE)

mevastatin_ft %>%
  join_metadata(mevastatin_metadata) %>%
  filter_grouped_mv(min_found = 0.75,
                    group_column = Group) %>%
  filter_blank(blank_samples = "blank", 
               blank_as_group = T,
               group_column = Group) %>%
  filter_cv(reference_samples = "QC",
            ref_as_group = T,
            group_column = Group) %>%
  dplyr::filter(Group != "blank") %>%
  impute_knn() %>% 
  normalize_quantile_smooth() %>%
  collapse_mean(sample_metadata_cols = "Group",
                feature_metadata_cols = "Feature") %>%
  plot_volcano(group_column = Group,
               name_column = Feature, 
               groups_to_compare = c("control", "mevastatin")) +
    ggplot2::geom_hline(yintercept = -log10(0.05),
                        color = "grey40",
                        linetype = 2) +
    ggplot2::geom_vline(xintercept = c(- 1, 1),
                        color = "grey40",
                        linetype = 2) +
    ggplot2::theme_bw()
```

## Read the required files

This vignette uses data from a separate [GitHub repository](https://github.com/yasche/metamorphr-data). More information about the data can be found there.

### Read the feature table

The first step in the workflow is to read a feature table. The metamorphr package contains a convenience function, `read_featuretable`, that reads files containing delimiter separated values, for example files in the csv format, and performs several initial tidying steps to bring them in the correct format. `read_featuretable` can read files from disk but also literal data and data from the internet. We use the latter to download data from a [GitHub repository](https://github.com/yasche/metamorphr-data).

```{r read-ft}
mevastatin_ft <- read_featuretable(
  "https://raw.githubusercontent.com/yasche/metamorphr-data/main/RP18/pos/MetaboScape/mevastatin/mevastatin.csv", 
  label_col = 3, 
  metadata_cols = 1:15)
```

With the `label_col` argument, we specify that the values in column 3 in [mevastatin.csv](https://github.com/yasche/metamorphr-data/blob/main/RP18/pos/MetaboScape/mevastatin/mevastatin.csv) are used as feature labels (in this case the exact mass). Those values are stored in a column named `Feature`. Further, with the `metadata_cols` argument we specify that columns 1 to 15 contain feature metadata, for example retention time and collision cross section (CCS). Those are preserved but it is important to tell `read_featuretable` which columns contain metadata so the parsing and initial tidying work as expected.

In addition to the `Feature` column and the preserved metadata columns, several other columns are created automatically by `read_featuretable`:

-   `UID`: a column containing a unique identifier for each feature

-   `Sample`: a column containing the sample names

-   `Intensity`: the measured intensity (or area) for a specific `Sample` `Feature` combination.

```{r ft-head}
head(mevastatin_ft)
```

### Read the sample metadata

Sample metadata, for example group information or replicate number, is required for some of the functions. An empty metadata table with required columns and rows can be created from an existing feature table via `create_metadata_skeleton` and exported and edited by hand, for example in Microsoft Excel. After editing, the file can be read back into R using `readr::read_csv`.

```{r md-example, eval=FALSE}
mevastatin_ft %>%
  create_metadata_skeleton() %>%
  readr::write_csv(file = "some/path/menadione_metadata.csv")

mevastatin_metadata <- readr::read_csv("some/path/menadione_metadata.csv")
```

In this case, there is already a file containing the relevant metadata in the GitHub repository.

```{r read-metadata}
mevastatin_metadata <- readr::read_csv(
  "https://raw.githubusercontent.com/yasche/metamorphr-data/main/RP18/pos/MetaboScape/mevastatin/mevastatin_metadata.csv", 
  show_col_types = FALSE)
```

```{r show-metadata}
head(mevastatin_metadata)
```

The metadata table must include the following columns:

-   `Sample`: Contains the sample names.

-   `Group`: Contains group information.

-   `Replicate`: Replicate information. If multiple replicates of the same sample exist (e.g., the same sample was injected multiple times) they must have the same replicate number.

-   `Batch`: Samples belonging to the same batch must have the same batch number.

-   `Factor`: A sample specific factor, for example protein concentration or dry weight. Only necessary for specific normalization strategies.

### Join them together

metamorphr provides a convenience function, `join_metadata`, for joining together a feature table and associated metadata.

```{r join-metadata}
mevastatin_ft <- join_metadata(mevastatin_ft, mevastatin_metadata)
```

## Filter

In this workflow, filter functions are used to improve the overall data quality. In the R console, you can type `metamorphr::filter_` to see all available filter functions. `dplyr::filter` is used to remove blank samples afterwards.

```{r filter-features}
mevastatin_ft <- mevastatin_ft %>%
  filter_grouped_mv(min_found = 0.75,
                    group_column = Group) %>%
  filter_blank(blank_samples = "blank", 
               blank_as_group = T,
               group_column = Group) %>%
  filter_cv(reference_samples = "QC",
            ref_as_group = T,
            group_column = Group) %>%
  dplyr::filter(Group != "blank")
```

In this workflow, 3 filter functions are used sequentially:

-   `filter_grouped_mv`: To only keep features with a low rate of missing values. With default parameters, only features are kept if they were found in at least 75 % of the samples of at least 1 group.

-   `filter_blank`: To remove features that were found with a high intensity in blank samples. With default parameters, only features are kept were the maximum intensity in non-blank samples is at least 3 times as high as in blank samples.

-   `filter_cv`: To remove features with a poor reproducibility (i.e., with a high coefficient of variation, CV) in quality control samples. With the default parameters, only features with a maximum CV of 0.2 are kept.

## Imputate missing values

The data contains missing values (i.e., `NA`). Those should be imputed prior to creating a volcano plot. For this example, the [*k*-nearest neighbors algorithm](https://en.wikipedia.org/wiki/K-nearest_neighbors_algorithm) is used for imputation but the metamorphr package provides other options as well. Start typing `metamorphr::impute_` in the R console to see the available options.

The underlying design of the metamorphr package makes it very easy to visualize the data using 'ggplot2'. Below, a simple visualization is used to verify that the imputation worked as expected.

```{r impute}
mevastatin_ft_before_impute <- mevastatin_ft %>%
  dplyr::mutate(State = "Before imputation")
  
mevastatin_ft <- impute_knn(mevastatin_ft)
```

```{r viz-impute}
mevastatin_ft %>%
  dplyr::mutate(State = "After imputation") %>%
  dplyr::bind_rows(mevastatin_ft_before_impute) %>%
  dplyr::mutate(State = factor(State, 
                               levels = c("Before imputation", 
                                          "After imputation"))) %>%
  dplyr::group_by(State, Sample, Group) %>%
  dplyr::count(wt = is.na(Intensity)) %>%
  ggplot2::ggplot(ggplot2::aes(Sample, n, color = Group)) +
    ggplot2::geom_point() +
    ggplot2::facet_wrap(~State, nrow = 2) +
    ggplot2::ylab("Number of missing values") +
    ggplot2::theme_bw() +
    ggplot2::theme(axis.text.x = ggplot2::element_blank(),
                   legend.position = "bottom")
```

As expected, there are no missing values after imputation.


## Normalize

Next, intensities are normalized across samples. In this example, a smooth quantile normalization is used (Hicks et al., 2018). This algorithm requires sample metadata, more specifically, grouping information. Other available options can be viewed by typing `metamorphr::normalize_` into the R console.

```{r normalize}
mevastatin_ft_before_norm <- mevastatin_ft %>%
  dplyr::mutate(State = "Before normalization")

mevastatin_ft <- normalize_quantile_smooth(mevastatin_ft)
```

Again, we can use a 'ggplot2' visualization as verification.


```{r viz-norm}
mevastatin_ft %>%
  dplyr::mutate(State = "After normalization") %>%
  dplyr::bind_rows(mevastatin_ft_before_norm) %>%
  dplyr::mutate(State = factor(State, 
                               levels = c("Before normalization", 
                                          "After normalization"))) %>%
  ggplot2::ggplot(ggplot2::aes(Sample, log10(Intensity), color = Group)) +
    ggplot2::geom_boxplot() +
    ggplot2::facet_wrap(~State, nrow = 2) +
    ggplot2::ylab("log10(Intensity)") +
    ggplot2::theme_bw() +
    ggplot2::theme(axis.text.x = ggplot2::element_blank(),
                   legend.position = "bottom")
```

The normalization worked as expected. The distributions of intensities inside each group are identical.

## Collapse technical replicates

In the example data set, the same sample was injected 3 times. Prior to doing statistics and drawing the volcano plot, those technical replicates should be collapsed. The metamorphr package provides several approaches to do this. Start typing `metamorphr::collapse_` into the R console to see the available options. For this workflow, we will simply calculate the mean of the technical replicates using `collapse_mean`. It is important that sample metadata are provided for this approach.

```{r show-sample-metadata}
mevastatin_metadata
```

It is important that `Group` and `Replicate` information is provided as well as `Batch`. Technical replicates of the same sample must have the same value for `Group`, `Batch` and `Replicate`. For example, `B9_Comp_1...` belongs to group `mevastatin` and was injected 3 times as indicated by the three samples with replicate number 1.


```{r}
mevastatin_ft <- collapse_mean(mevastatin_ft,
                               sample_metadata_cols = "Group",
                               feature_metadata_cols = "Feature")
```

## Plot

Finally, data can be plotted with `plot_volcano`. The resulting plot can be easily modified with 'ggplot2' functions.

```{r plot-volc, warning=FALSE}
plot_volcano(mevastatin_ft,
             group_column = Group,
             name_column = Feature, 
             groups_to_compare = c("control", "mevastatin")) +
  ggplot2::geom_hline(yintercept = -log10(0.05),
                      color = "grey40",
                      linetype = 2) +
  ggplot2::geom_vline(xintercept = c(- 1, 1),
                      color = "grey40",
                      linetype = 2) +
  ggplot2::theme_bw()
```

## References

Hicks, S. C., Okrah, K., Paulson, J. N., Quackenbush, J., Irizarry, R. A., & Bravo, H. C. (2018). Smooth quantile normalization. *Biostatistics, 19*(2), 185–198. <https://doi.org/10.1093/biostatistics/kxx028>
